Freezing of a Weeks-Chandler-Anderson Fluid using classical Density Functional Theory

### 2020-09-08 papers/thesis-kirstie/thesis.tex
a bit more simplified

The key idea behind classical Density Functional Theory (cDFT) is that free energy can be written as a functional (a function of a function) of a number density profile $n(\vec{r})$ which describes the way the number density, or atoms per volume, varies spatially. 
\begin{displaymath}{f[n(}\vec{r}{)]}\end{displaymath}

This, coupled with the fact that the free energy of a system tends toward a minimum at equilibrium means that the equilibrium free energy of the system ${f_{equil}}{[}\rho{(}\vec{r}{)]}$, and its corresponding equilibrium number density profile $\rho{(}\vec{r}{)}$, can be found simply by varying $n(\vec{r})$ through all possible spatial profiles until the free energy is minimized.

@@

The reason this works is because cDFT satisfies the conditions of the variational principle, the same mathematical principle used to determine that the shortest distance between two points is a straight line by varying the path between the two points until the distance is minimized.

A description of cDFT ordinarily begins with considering the Grand Free Energy defined as
\begin{equation}\Phi=F-\mu{N}\end{equation}
%\begin{equation}\int\frac{N}{V}\frac{V_{ext}}{N}dV\end{equation}
which can then be expressed in terms of a varrying number density profile $n(\vec r)$
%\begin{equation}\Phi=F-\int n(\vec{r})\phi{(\vec r)}d\vec{r} +\int n(\vec{r})\phi{(\vec r)}d\vec{r}-\mu{N}\end{equation}
\begin{equation}\label{GrandFE}\Phi= F_{int} +\int n(\vec{r})\phi{(\vec r)}d\vec{r}-\mu\int n(\vec r)d\vec{r}\end{equation}
where $\phi=\frac{V_{ext}}{N}$, $V_{ext}(\vec r)$ is an externally applied potential energy, and
\begin{equation}F_{int}= F -\int n(\vec{r})\phi{(\vec r)}d\vec{r}\end{equation}  
%If $\mu=0$ and $V_{ext}=0$, then the Grand free energy $\Phi$ reduces to the Helmholz free energy $F$. %The second and third terms in Eq~\ref{GrandFE} are both functionals of $n(\vec{r})$, so it remains to show that $F$, and thus $\Phi$, is a functional of $n(\vec{r})$.

The Grand free energy is applicable to systems with a varying number of particles and fixed chemical potential, temperature and volume. But this paper will be examining a system with a fixed temperature, volume, and number of particles in which case it is the Helmholtz energy as a functional of the number density profile $F[n(\vec r)]$ that is of interest.

The Helmholtz free energy F is defined as 
\begin{equation}F=U-TS\end{equation}
\begin{equation}\frac{F}{N}=\frac{U}{N}-\frac{TS}{N}\end{equation}
\begin{equation}\label{usetoshowmin}d\left(\frac{F}{N}\right)=d\left(\frac{U}{N}\right)-d\left(\frac{TS}{N}\right)\end{equation}
\begin{equation}d\left(\frac{F}{N}\right)=d\left(\frac{U}{N}\right)-\frac{T}{N}dS-\frac{S}{N}dT-TS d\left(\frac{1}{N}\right)\end{equation}
Setting $f_N=\frac{F}{N}$ this becomes
\begin{equation}df_N=-\frac{S}{N}dT-\frac{P}{N}dV-PVd\left(\frac{1}{N}\right)\end{equation}
This result shows that the natural variables of the Helmholtz free energy per particle $f_N$ are temperature, volume and one over the number of particles, which in turn also holds the number density $n=\frac{V}{N}$ fixed. It is the Helmholtz free energy will be minimized in a system that has a temperature, volume, and number of particles that are fixed.
%\begin{equation}f_N(T,V, \frac{1}{N})\mbox{~~~~or~~~~}f_N(T,n)\end{equation}
\begin{equation}f_N(T,V, \frac{1}{N})\end{equation}

@@

At low densities, it is not likely to find two or more atoms in close proximity, and so only interactions between individual pairs of spheres, such as those shown in green in Figure~\ref{fig:diagrammic}, are considered.
Each pair can then be described by a function of the distance between the two atoms in the pair. Making a change of variables $\vec{r}=\vec{r}_i-\vec{r}_j $ for each pair we get 
\begin{equation}f_{ij}(\vec{r_i}-\vec{r_j})=f(\vec{r})=e^{-\frac{V(\vec{r})}{k_BT}}-1\end{equation} 
\color{black}
With a careful counting of the resulting number of such pairs, the excess free energy can then be written as 
where in the thermodynamic limit (large N), $N-1, N-2, ...\approx{N}$. 
Written in terms of an exponential $e^x=1+x+\frac{x^2}{2!}+\frac{x^3}{3!}+ ...$ for a homogeneous density n=$\frac{N}{V}$ this becomes
\begin{equation}{F_{excess}=-k_BT\ln{\left(e^{\frac{1}{2}\frac{N^2}{V}\int{f(r)}{d\vec{r}}}\right)}}\end{equation} 
Thus the free excess energy per atom in the low-density limit is given by 
\begin{equation}{\frac{F_{excess}}{N}=k_BTn\left(-\frac{1}{2}\int{f(r)}{d\vec{r}}\right)}\end{equation} where the quantity in parenthesis can be recognized 
as the second virial expanision coefficient typically called $B_2$. %where $x=\frac{1}{2}\frac{N(N-1)}{V^{2}}\int{...}\int{dQ}{f_{12}}$.% 
It is important that the expression for the excess Helmholtz free energy correctly reduces to this result in the low-density limit. 
%Although we will not be restricting ourselves to the low-density limit, it is important that the theory we use correctly reduces to this same result in the low-density limit. 
Outside of the low density limit, it is quite difficult to evaluate the excess Helmholtz free energy given in Equation~\ref{Fexcess-simplified}. 
Another way to form an expression for the excess Helmholtz free energy in terms of a homogeneous number density is to use Scaled Particle Theory (SPT).

>>>>

The key idea behind classical Density Functional Theory (cDFT) is that free energy can be written as a functional (a function of a function) of a number density profile $n(\vec{r})$ which describes the way the number density, or atoms per volume, varies spatially. 
\begin{displaymath}{f[n(}\vec{r}{)]}\end{displaymath}
This, coupled with the fact that the free energy of a system tends toward a minimum at equilibrium means that the equilibrium free energy of the system ${f_{equil}}{[}\rho{(}\vec{r}{)]}$, and its corresponding equilibrium number density profile $\rho{(}\vec{r}{)}$, can be found simply by varying $n(\vec{r})$ through all possible spatial profiles until the free energy is minimized.

@@

%The reason this works is because cDFT satisfies the conditions of the variational principle, the same mathematical principle used to determine that the shortest distance between two points is a straight line by varying the path between the two points until the distance is minimized.

In this paper a system with a fixed temperature, volume, and number of particles will be examined in which case it is the Helmholtz energy as a functional of the number density profile $F[n(\vec r)]$ that is of interest. The Helmholtz free energy F is defined as 
\begin{equation}F=U-TS\end{equation}
\begin{equation}\frac{F}{N}=\frac{U}{N}-\frac{TS}{N}\end{equation}
\begin{equation}\label{usetoshowmin}d\left(\frac{F}{N}\right)=d\left(\frac{U}{N}\right)-d\left(\frac{TS}{N}\right)\end{equation}
\begin{equation}d\left(\frac{F}{N}\right)=d\left(\frac{U}{N}\right)-\frac{T}{N}dS-\frac{S}{N}dT-TS d\left(\frac{1}{N}\right)\end{equation}
Setting $f_N=\frac{F}{N}$ this becomes
\begin{equation}df_N=-\frac{S}{N}dT-\frac{P}{N}dV-PVd\left(\frac{1}{N}\right)\end{equation}
This result shows that the natural variables of the Helmholtz free energy per particle $f_N$ are temperature, volume and one over the number of particles, which in turn also holds the number density $n=\frac{V}{N}$ fixed. It is the Helmholtz free energy will be minimized in a system that has a temperature, volume, and number of particles that are fixed.
%\begin{equation}f_N(T,V, \frac{1}{N})\mbox{~~~~or~~~~}f_N(T,n)\end{equation}
\begin{equation}f_N(T,V, \frac{1}{N})\end{equation}

@@

At low densities, it is not likely to find two or more atoms in close proximity, and so only interactions between individual pairs of spheres, such as those shown in green in Figure~\ref{fig:diagrammic}, are considered.
Each pair can then be described by a function of the distance between the two atoms in the pair. Making a change of variables $\vec{r}=\vec{r}_i-\vec{r}_j $ for each pair we get 
\begin{equation}f_{ij}(\vec{r_i}-\vec{r_j})=f(\vec{r})=e^{-\frac{V(\vec{r})}{k_BT}}-1\end{equation} 
\color{black}
With a careful counting of the resulting number of such pairs, the excess free energy can then be written as 
\begin{equation}{F_{excess}=-k_BT\ln\left(1+\frac{1}{2}\frac{N^2}{V}\int{f(r)}{d\vec{r}}+\frac{1}{2}\left(\frac{1}{2}\frac{N^2}{V}\int{f(r)}{d\vec{r}}\right)^2+ ...\right)}\end{equation}
where in the thermodynamic limit (large N), $N-1, N-2, ...\approx{N}$. 
Written in terms of an exponential $e^x=1+x+\frac{x^2}{2!}+\frac{x^3}{3!}+ ...$ for a homogeneous density n=$\frac{N}{V}$ this becomes
\begin{equation}{F_{excess}=-k_BT\ln{\left(e^{\frac{1}{2}\frac{N^2}{V}\int{f(r)}{d\vec{r}}}\right)}}\end{equation} 
Thus the free excess energy per atom in the low-density limit is given by 
\begin{equation}{\frac{F_{excess}}{N}=k_BTn\left(-\frac{1}{2}\int{f(r)}{d\vec{r}}\right)}\end{equation} where the quantity in parenthesis can be recognized 
as the second virial expanision coefficient typically called $B_2$. %where $x=\frac{1}{2}\frac{N(N-1)}{V^{2}}\int{...}\int{dQ}{f_{12}}$.% 
It is important that the expression for the excess Helmholtz free energy correctly reduces to this result in the low-density limit. 
%Although we will not be restricting ourselves to the low-density limit, it is important that the theory we use correctly reduces to this same result in the low-density limit. 
Outside of the low density limit, it is quite difficult to evaluate the excess Helmholtz free energy given in Equation~\ref{Fexcess-simplified}. 
Another way to form an expression for the excess Helmholtz free energy in terms of a homogeneous number density is to use Scaled Particle Theory (SPT).
